Deep mutational scanning of hemagglutinin helps predict evolutionary fates of human H3N2 influenza variants

### 2018-04-06 paper/manuscript.tex
Updated 1968-2018 phylogeny.
Update figures, incorporate into manuscript

\begin{figure}[h!]
\centering
\includegraphics[width=\linewidth]{figs/frequency_trajectory_ex.pdf}
\caption{\label{fig:frequency_trajectory}
{\bf Frequency trajectories of individual mutations and their relation to the experimentally determined effects of these mutations.}
The top panel shows the subset of the full H3N2 HA tree in Figure~\ref{suppfig:tree} windowed from 2004 to 2014.
Circles indicate individual amino-acid mutations, and are colored according to the mutational effect measured in our deep mutational scanning (negative values indicate mutations measured to be deleterious to viral growth).
The bottom panel shows the frequency trajectory of each mutation, with mutation trajectories colored according to the mutational effects as in the top panel.
It is clear that most mutations that reach high frequency are measured to be relatively favorable in our experiments.
}
\end{figure}

\begin{figure*}[ht]
\centering
\includegraphics[width=0.85\textwidth]{figs/muteffect_by_maxfreq.pdf}
\caption{\label{fig:muteffect_maxfreq}
{\bf Experimental measurements are informative about the evolutionary fate of viral mutations.}
(A) Correlation between the effects of mutations as measured in our experiments and the maximum frequency reached by these mutations.
The plots show Spearman $\rho$ and an empirical $P$-value representing the proportion of 10,000 permutations for which permuted $\rho$ was greater than or equal to the observed $\rho$.
(B) The distribution of mutational effects partitioned by maximum mutation frequency.
The vertical black line shows the mean mutation effect for each category.
The analysis is performed separately for the pre- and post-Perth/2009 partitions of the tree (see Figure~\ref{suppfig:tree}).
}
\end{figure*}

@@

\begin{figure*}[ht]
\centering
\includegraphics[width=1.0\textwidth]{figs/muteffect_by_maxfreq_WSN_1.pdf}
\caption{\label{fig:muteffect_maxfreq_WSN}
{\bf Experimental measurements on an H1 HA are not informative about the evolutionary fate of H3N2 mutations.}
This figure repeats the analysis of the H3N2 mutation frequencies in Figure~\ref{fig:muteffect_maxfreq}A, but uses the deep mutational scanning data for an H1 HA as measured in \cite{doud2016accurate}, i.e.\
Figure~\ref{suppfig:muteffect_maxfreq_WSN_supp} shows the histograms comparable to those in Figure~\ref{fig:muteffect_maxfreq}B.
The empirical $P$-value represents the results of 1,000 permutations.
}
\end{figure*}

@@

\begin{suppfigure}[H]
\centerline{\includegraphics[width=0.5\textwidth]{figs/S01_G78D-T212I/G78D-T212I.pdf}}
\caption{\label{suppfig:Perth2009_mut}
{\bf Characterization of the G78D-T212I Perth/2009 HA variant.}
(A)
The G78D-T212I Perth/2009 HA variant supports better viral growth than the wildtype Perth/2009 HA.
Viruses were generated in duplicate by reverse genetics with the Perth/2009 NA and WSN internal genes, and passaged once at MOI = 0.01 in MDCK-SIAT1-TMPRSS2 cells.
The rescue and passage viral supernatants were collected at 72 hours post-transfection and 44 hours post-infection, respectively, and titered in MDCK-SIAT1-TMPRSS2 cells.
The points mark each duplicate and the bar marks the mean.
(B)
The D78 variant remained at a low frequency in natural human H3N2 sequences over the past $~\sim$10 years.
The A212 variant rose to fixation in $~\sim$2011, replacing the T212 variant.
}
\end{suppfigure}

\begin{suppfigure}[H]
\centerline{\includegraphics[width=0.6\textwidth]{figs/S02_SangerSeq/SangerSeq.pdf}}
\caption{\label{suppfig:SangerSeq}
{\bf Sanger sequencing of 31 randomly chosen clones from the mutant plasmid libraries.}
(A) There were an average of $\sim$1.4 codon mutations per clone across the three plasmid mutant libraries.
(B) A mixture of one-, two-, and three-nucleotide mutations were present, with slightly fewer triple-nucleotide changes than expected.
(C) Nucleotide frequencies were uniform in the codon mutations.
(D) The mutations were distributed relatively evenly across the length of the HA coding sequence.
(E) We calculated the pairwise distances between mutations for clones carrying more than one mutation.
The cumulative distribution of these distances is shown in the red line.
The blue line indicates the expected distribution if mutations in multiply mutated genes are randomly dispersed along the sequence.
}
\end{suppfigure}

\begin{suppfigure}[H]
\centerline{\includegraphics[width=\textwidth]{figs/S03_WSNprefs_logoplot/WSN-rescaled_prefs.pdf}}
\caption{\label{suppfig:WSNprefs_logoplot}
{\bf The site-specific amino-acid preferences of the WSN/1933 H1 HA as measured in \cite{doud2016accurate}.}
The amino-acid preferences from~\cite{doud2016accurate} after taking the average of the experimental replicates and re-scaling~\cite{hilton2017phydms} by a stringency parameter of 2.05 (see \url{https://github.com/jbloomlab/dms_tools2/blob/master/examples/Doud2016/analysis_notebook.ipynb}).
The sites are in H3 numbering.
The overlays show the same information as in Figure~\ref{fig:logoplot} (domain and wildtype amino acid).
}
\end{suppfigure}

\begin{suppfigure}[H]
\centerline{\includegraphics[clip=true, trim=0 5.5in 0 0,width=0.7\textwidth]{figs/S04_post_Perth2009/post_Perth2009_seqpref.pdf}}
\caption{\label{suppfig:tree}
{\bf A phylogenetic tree of all HA sequences used in our analysis of mutation frequencies.}
HA sequences were sampled at a rate of six viruses per month from January 1, 1968 through February 1, 2018.
The Perth/2009 strain used in our experiments is indicated.
The rest of the tree is partitioned into nodes that preceded the split of the Perth/2009 strain from the trunk of the tree (blue) and nodes that branched off the trunk after the clade containing Perth/2009 (orange).
In Figure~\ref{fig:muteffect_maxfreq}, these two partitions of the tree are analyzed separately.
Nodes in the clade containing the Perth/2009 strain and nodes sampled in 2014 or after were excluded from our analyses.
}
\end{suppfigure}

@@

\begin{suppfigure}[H]
\centerline{\includegraphics[width=1.0\textwidth]{figs/muteffect_by_maxfreq_WSN_2.pdf}}
\caption{\label{suppfig:muteffect_maxfreq_WSN_supp}
{\bf The distribution of mutational effects measured in H1 HA among H3N2 mutations binned by the maximum frequency that they reach.}
This figure repeats the analysis of the H3N2 mutation frequencies in Figure~\ref{fig:muteffect_maxfreq}B, but uses the deep mutational scanning data for an H1 HA as measured in \cite{doud2016accurate}.
}
\end{suppfigure}

@@

>>>>

\begin{figure}[h!]
\centering
\includegraphics[width=\linewidth]{figs/frequency_trajectory_ex/frequency_trajectory_ex.pdf}
\caption{\label{fig:frequency_trajectory}
{\bf Frequency trajectories of individual mutations and their relation to the experimentally determined effects of these mutations.}
The top panel shows the subset of the full H3N2 HA tree in Figure~\ref{suppfig:tree} windowed from 2004 to 2014.
Circles indicate individual amino-acid mutations, and are colored according to the mutational effect measured in our deep mutational scanning (negative values indicate mutations measured to be deleterious to viral growth).
The bottom panel shows the frequency trajectory of each mutation, with mutation trajectories colored according to the mutational effects as in the top panel.
It is clear that most mutations that reach high frequency are measured to be relatively favorable in our experiments.
}
\end{figure}

\begin{SCfigure*}[\sidecaptionrelwidth][t]
%\begin{figure*}[ht]
\centering
\includegraphics[width=12cm]{figs/muteffect_by_maxfreq/muteffect_by_maxfreq.pdf}
%\includegraphics[width=0.85\textwidth]{figs/muteffect_by_maxfreq/muteffect_by_maxfreq.pdf}
\caption{\label{fig:muteffect_maxfreq}
{\bf Experimental measurements are informative about the evolutionary fate of viral mutations.}
(A) Correlation between the effects of mutations as measured in our experiments and the maximum frequency reached by these mutations.
The plots show Spearman $\rho$ and an empirical $P$-value representing the proportion of 10,000 permutations for which permuted $\rho$ was greater than or equal to the observed $\rho$.
(B) The distribution of mutational effects partitioned by maximum mutation frequency.
The vertical black line shows the mean mutation effect for each category.
The analysis is performed separately for the pre- and post-Perth/2009 partitions of the tree (see Figure~\ref{suppfig:tree}).
}
\end{SCfigure*}
%\end{figure*}

@@

\begin{SCfigure*}[\sidecaptionrelwidth][t]
%\begin{figure*}[ht]
\centering
\includegraphics[width=12cm]{figs/muteffect_by_maxfreq_WSN/muteffect_by_maxfreq_WSN.pdf}
%\includegraphics[width=0.85\textwidth]{figs/muteffect_by_maxfreq_WSN/muteffect_by_maxfreq_WSN.pdf}
\caption{\label{fig:muteffect_maxfreq_WSN}
{\bf Experimental measurements on an H1 HA are not informative about the evolutionary fate of H3N2 mutations.}
This figure repeats the analysis of the H3N2 mutation frequencies in Figure~\ref{fig:muteffect_maxfreq}A, but uses the deep mutational scanning data for an H1 HA as measured in \cite{doud2016accurate}, i.e.\
Figure~\ref{suppfig:muteffect_maxfreq_WSN_supp} shows the histograms comparable to those in Figure~\ref{fig:muteffect_maxfreq}B.
The empirical $P$-value represents the results of 1,000 permutations.
}
\end{SCfigure*}
%\end{figure*}

@@

\begin{suppfigure}[H]
\centerline{\includegraphics[width=0.5\textwidth]{figs/supp_G78D-T212I/G78D-T212I.pdf}}
\caption{\label{suppfig:Perth2009_mut}
{\bf Characterization of the G78D-T212I Perth/2009 HA variant.}
(A)
The G78D-T212I Perth/2009 HA variant supports better viral growth than the wildtype Perth/2009 HA.
Viruses were generated in duplicate by reverse genetics with the Perth/2009 NA and WSN internal genes, and passaged once at MOI = 0.01 in MDCK-SIAT1-TMPRSS2 cells.
The rescue and passage viral supernatants were collected at 72 hours post-transfection and 44 hours post-infection, respectively, and titered in MDCK-SIAT1-TMPRSS2 cells.
The points mark each duplicate and the bar marks the mean.
(B)
The D78 variant remained at a low frequency in natural human H3N2 sequences over the past $~\sim$10 years.
The A212 variant rose to fixation in $~\sim$2011, replacing the T212 variant.
}
\end{suppfigure}

\begin{suppfigure}[H]
\centerline{\includegraphics[width=0.6\textwidth]{figs/supp_SangerSeq/SangerSeq.pdf}}
\caption{\label{suppfig:SangerSeq}
{\bf Sanger sequencing of 31 randomly chosen clones from the mutant plasmid libraries.}
(A) There were an average of $\sim$1.4 codon mutations per clone across the three plasmid mutant libraries.
(B) A mixture of one-, two-, and three-nucleotide mutations were present, with slightly fewer triple-nucleotide changes than expected.
(C) Nucleotide frequencies were uniform in the codon mutations.
(D) The mutations were distributed relatively evenly across the length of the HA coding sequence.
(E) We calculated the pairwise distances between mutations for clones carrying more than one mutation.
The cumulative distribution of these distances is shown in the red line.
The blue line indicates the expected distribution if mutations in multiply mutated genes are randomly dispersed along the sequence.
}
\end{suppfigure}

\begin{suppfigure}[H]
\centerline{\includegraphics[width=\textwidth]{figs/supp_WSNprefs_logoplot/WSN-rescaled_prefs.pdf}}
\caption{\label{suppfig:WSNprefs_logoplot}
{\bf The site-specific amino-acid preferences of the WSN/1933 H1 HA as measured in \cite{doud2016accurate}.}
The amino-acid preferences from~\cite{doud2016accurate} after taking the average of the experimental replicates and re-scaling~\cite{hilton2017phydms} by a stringency parameter of 2.05 (see \url{https://github.com/jbloomlab/dms_tools2/blob/master/examples/Doud2016/analysis_notebook.ipynb}).
The sites are in H3 numbering.
The overlays show the same information as in Figure~\ref{fig:logoplot} (domain and wildtype amino acid).
}
\end{suppfigure}

\begin{suppfigure}[H]
%%\centerline{\includegraphics[clip=true, trim=0 5.5in 0 0,width=0.7\textwidth]{figs/S04_post_Perth2009/post_Perth2009_seqpref.pdf}}
\centerline{\includegraphics[width=1.0\textwidth]{figs/supp_H3N2_phylogeny/H3N2_phylogeny.pdf}}
\caption{\label{suppfig:tree}
{\bf A phylogenetic tree of all HA sequences used in our analysis of mutation frequencies.}
HA sequences were sampled at a rate of six viruses per month from January 1, 1968 through February 1, 2018.
The Perth/2009 strain used in our experiments is indicated.
The rest of the tree is partitioned into nodes that preceded the split of the Perth/2009 strain from the trunk of the tree (blue) and nodes that branched off the trunk after the clade containing Perth/2009 (orange).
In Figure~\ref{fig:muteffect_maxfreq}, these two partitions of the tree are analyzed separately.
Nodes in the clade containing the Perth/2009 strain and nodes sampled in 2014 or after were excluded from our analyses.
}
\end{suppfigure}

@@

\begin{suppfigure}[H]
\centerline{\includegraphics[width=0.8\textwidth]{figs/supp_muteffect_maxfreq_WSN/muteffect_maxfreq_WSN_supp.pdf}}
\caption{\label{suppfig:muteffect_maxfreq_WSN_supp}
{\bf The distribution of mutational effects measured in H1 HA among H3N2 mutations binned by the maximum frequency that they reach.}
This figure repeats the analysis of the H3N2 mutation frequencies in Figure~\ref{fig:muteffect_maxfreq}B, but uses the deep mutational scanning data for an H1 HA as measured in \cite{doud2016accurate}.
}
\end{suppfigure}

@@

\begin{suppdata}[H]
\caption{\label{suppdata:gisaid}
Excel file providing acknowledgments and accessions for sequences downloaded from GISAID.
}
\end{suppdata}
